Formal Analysis of EDHOC Key Establishment for Constrained IoT Devices

### 2022-07-21 formalization.tex
Minor changes, fixing typos etc

% !TEX root = paper.tex
\vnote{Need to rewrite this based on the new spec}
The \mEdhoc{} \mSpec{} \cite{our-analysis-selander-lake-edhoc-00} claims
that \mEdhoc{} satisfies many security properties, but these are imprecisely
expressed and motivated.
%
In particular, there is no coherent adversary model.
%
It is therefore not clear in which context properties should be verified.
%
We resolve this by clearly specifying an adversary model, in which we can verify
properties.
%

\subsection{Formal Model}\label{sec:threat-model}
\vnote{Can we add extra references here? SAPIC+ etc? Can leak session keys as well.}

As in~\cite{Norr21}, we verify \mEdhoc{} in an extended
Dolev-Yao model~\cite{DY83}.
%
Dolev-Yao is a well-established model for symbolic verification of security
protocols.
%
Messages are modelled as terms in an algebra, and the various cryptographic
operations are assumed to be perfect, e.g., encrypted messages can only be
decrypted with the correct key, and there are no hash collisions exist.
%

@@

In this work, we consider the following properties:
perfect forward secrecy, injective agreement for both the initiator and the
responder, and implicit agreement.
%
Agreement is considered on a set of parameters $S$ which also contains the
session key material \mSessKey{}.
%
We will first describe in detail both properties, and then describe the
contents of the set $S$.
%
We formalize these properties as shown in Figure~\ref{fig:props}, which is
a modified version of a figure in~\cite{Norr21}.
\vnote{The figure changes because of TEEs}

\begin{figure*}[ht]
\begin{align*}
    \mPredPfs \triangleq\ & \forall I, R, \mSessKey, t_2, t_3\mLogicDot
    \mK^{t_3}(\mSessKey)\  \land\ 
    (\mIComplete^{t_2}(I, R, \mSessKey)\, \lor\, \mRComplete^{t_2}(I, R, \mSessKey))
    \rightarrow\\
    &(\exists t_1\mLogicDot \mRevLTK^{t_1}(I) \land t_1 \lessdot t_2)
    \ \lor\ (\exists t_1\mLogicDot \mRevLTK^{t_1}(R) \land t_1 \lessdot t_2)
    \ \lor\ (\exists t_1\mLogicDot \mRevEph^{t_1}(R, \mSessKey))
    \ \lor\ (\exists t_1\mLogicDot \mRevEph^{t_1}(I, \mSessKey))
\end{align*}
\begin{align*}
    \mPredInjI \triangleq\ &
    \forall I, R, \mSessKey, S, t_2\mLogicDot \mIComplete^{t_2}(I, R, \mSessKey, S)
    \rightarrow\\
    &(\exists t_1\mLogicDot \mRStart^{t_1}(R, \mSessKey, S) \land t_1 \lessdot t_2)
    \land (\forall I' R' t_1' \mLogicDot \mIComplete^{t_1'}(I' , R', \mSessKey, S)
        \rightarrow t_1' \doteq t_1)
    \ \ \lor\ \ (\exists t_1\mLogicDot \mRevLTK^{t_1}(R) \land t_1 \lessdot t_2)
\end{align*}
\begin{align*}
    \mPredInjR \triangleq\ &
    \forall I, R, \mSessKey, S, t_2\mLogicDot \mRComplete^{t_2}(I, R, \mSessKey, S)
    \rightarrow\\
    &(\exists t_1\mLogicDot \mIStart^{t_1}(I, R, \mSessKey, S) \land t_1 \lessdot t_2)
    \land (\forall I' R' t_1' \mLogicDot \mRComplete^{t_1'}(I' , R', \mSessKey, S)
        \rightarrow t_1' \doteq t_1)
    \ \ \lor\ \ (\exists t_1\mLogicDot \mRevLTK^{t_1}(I) \land t_1 \lessdot t_2).
\end{align*}
\begin{align*}
    \mPredImpI \triangleq\ &
    \forall I, R, \mSessKey, S, t_1\mLogicDot \mIComplete^{t_1}(I, R, \mSessKey, S)
    \rightarrow\\
      &(\forall I', R', S', t_2\mLogicDot \mRComplete^{t_2}(I', R', \mSessKey, S') \rightarrow
             (I=I' \land R=R' \land S=S'))\\
      &\land (\forall I', R', S', t_1'\mLogicDot
        (\mIComplete^{t_1'}(I', R', \mSessKey, S') \rightarrow t_1' \doteq t_1
        )\\
      &\ \ \ \lor(\exists t_0\mLogicDot \mRevLTK^{t_0}(R) \land t_0 \lessdot t_1)
    \lor(\exists t_0\mLogicDot \mRevEph^{t_0}(R, \mSessKey))
    \lor(\exists t_0\mLogicDot \mRevEph^{t_0}(I, \mSessKey)).
\end{align*}
%
\caption{Formalization of security properties and adversary model.}
\label{fig:props}
\end{figure*}

@@

Having described how we adapt the equational theory to model \mEdhoc,
we now move on to the modelling of the adversary model and the environment in
which the protocol is executed.
%
We extend the built-in Dolev-Yao adversary rules which are part of \mTamarin.
%
We use the following rules to capture the link between a party's identity and
their long-term key pairs, in the \mSig{}- and in the \mStat{}-based methods
respectively.
\begin{center}
\begin{minipage}{0.48\textwidth}
\begin{small}
\begin{verbatim}
1 rule registerLTK_SIG:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_SIG($A, ~ltk),
4         !PK_SIG($A, pk(~ltk)),
5         Out(<$A, pk(~ltk)>)]
\end{verbatim}
\end{small}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}{0.48\textwidth}
\begin{small}
\begin{verbatim}
1 rule registerLTK_STAT:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_STAT($A, ~ltk),
4         !PK_STAT($A, 'g'^~ltk),
5         Out(<$A, 'g'^~ltk>)]
\end{verbatim}
\end{small}
\end{minipage}
\end{center}

@@

As we extended the model with TEEs and augmented the adversary's capabilities
with access to them, \mTamarin{} failed to complete in a reasonable time for
some combination of authentication methods and security properties (see XYZ)
\knote{add reference to discussion section on this}.
%
To circumvent the problem, we simplified the \mXor{} encryption to \mXor-ing a
single term on the entire tuple for yjese cases.
%

{
\bgroup
\parindent 0pt
\begin{minipage}{0.49\textwidth}
\bgroup
\parindent 0pt
\begin{scriptsize}
\begin{verbatim}
1 rule R2_STAT_SIG:
2 let
3    agreed = <CS0, CI, ~CR>
4    gx = `g'^xx
5    data_2 = <`g'^~yy, CI, ~CR>
6    m1 = <`STAT', `SIG', CS0, CI, gx>
7    TH_2 = h(<$H0, m1, data_2>)
8    prk_2e = extr(`e', gx^~yy)
9    prk_3e2m = prk_2e
10   K_2m = expa(<$cAEAD0, TH_2, `K_2m'>,
11               prk_3e2m)
12   protected2 = $V // ID_CRED_V
13   CRED_V = pkV
14   extAad2 = <TH_2, CRED_V>
15   assocData2 = <protected2, extAad2>
16   MAC_2 = aead(`e', K_2m, assocData2,
                  $cAEAD0)
17   authV = sign(<assocData2, MAC_2>, ~ltk)
18   plainText2 = <$V, authV>
\end{verbatim}
\end{scriptsize}
\egroup
\end{minipage}
\begin{minipage}{0.02\textwidth}
\end{minipage}
\begin{minipage}{0.48\textwidth}
\bgroup
\parindent 0pt
\begin{scriptsize}
\begin{verbatim}
19   K_2e = expa(<$cAEAD0, TH_2,
                 `K_2e'>, prk_2e)
20   K_2e_1 = expa(<$cAEAD0, TH_2,
                   `K_2e', `1'>, prk_2e)
21   K_2e_2 = expa(<$cAEAD0, TH_2,
                   `K_2e', `2'>, prk_2e)
22   CIPHERTEXT_2 = <$V XOR K_2e_1,
                     authV XOR K_2e_2>
23   m2 = <data_2, CIPHERTEXT_2>
24   exp_sk = <gx^~yy>
25 in
26   [!LTK_SIG($V, ~ltk), !PK_SIG($V, pkV),
      In(m1), Fr(~CR), Fr(~yy), Fr(~tid)]
27   --[ExpRunningR(~tid, $V, exp_sk, agreed),
        R2(~tid, $V, m1, m2)]->
28   [StR2_STAT_SIG($V, ~ltk, ~yy, prk_3e2m,
              TH_2, CIPHERTEXT_2, gx^~yy,
              ~tid, m1, m2, agreed),
29    Out(m2)]
\end{verbatim}
\end{scriptsize}
\egroup
\end{minipage}
\\
\egroup
%
}

@@

The properties we listed in Section~\ref{sec:desired-properties}
translate directly into \mTamarin's logic.
%
We show the \mTamarin{} lemma which encodes the \mPredPfs{} property.
%
Other properties are formalized similarly.
%

>>>>

% !TEX root = paper.tex
The \mEdhoc{} \mSpec{} \cite{our-analysis-selander-lake-edhoc-00} claims
that \mEdhoc{} satisfies many security properties, but these are imprecisely
expressed and motivated.
%
In particular, there is no coherent adversary model.
%
It is therefore not clear in which context properties should be verified.
%
We resolve this by clearly specifying an adversary model, in which we can verify
properties.
%

\subsection{Formal Model}\label{sec:threat-model}
As in~\cite{Norr21}, we verify \mEdhoc{} in an extended
Dolev-Yao model~\cite{DY83}.
%
Dolev-Yao is a well-established model for symbolic verification of security
protocols.
%
Messages are modelled as terms in an algebra, and the various cryptographic
operations are assumed to be perfect, e.g., encrypted messages can only be
decrypted with the correct key, and there are no hash collisions exist.
%

@@

In this work, we consider the following properties:
perfect forward secrecy, injective agreement for both the initiator and the
responder, and implicit agreement.
%
Agreement is considered on a set of parameters $S$ which also contains the
session key material \mSessKey{}.
%
We will first describe in detail both properties, and then describe the
contents of the set $S$.
%
We formalize these properties as shown in Figure~\ref{fig:props}, which is
a modified version of a figure in~\cite{Norr21}.
\begin{figure*}[ht]
\begin{align*}
    \mPredPfs \triangleq\ & \forall I, R, \mSessKey, t_2, t_3\mLogicDot
    \mK^{t_3}(\mSessKey)\  \land\ 
    (\mIComplete^{t_2}(I, R, \mSessKey)\, \lor\, \mRComplete^{t_2}(I, R, \mSessKey))
    \rightarrow\\
    &(\exists t_1\mLogicDot \mRevLTK^{t_1}(I) \land t_1 \lessdot t_2)
    \ \lor\ (\exists t_1\mLogicDot \mRevLTK^{t_1}(R) \land t_1 \lessdot t_2)
    \ \lor\ (\exists t_1\mLogicDot \mRevEph^{t_1}(R, \mSessKey))
    \ \lor\ (\exists t_1\mLogicDot \mRevEph^{t_1}(I, \mSessKey))
\end{align*}
\begin{align*}
    \mPredInjI \triangleq\ &
    \forall I, R, \mSessKey, S, t_2\mLogicDot \mIComplete^{t_2}(I, R, \mSessKey, S)
    \rightarrow\\
    &(\exists t_1\mLogicDot \mRStart^{t_1}(R, \mSessKey, S) \land t_1 \lessdot t_2)
    \land (\forall I' R' t_1' \mLogicDot \mIComplete^{t_1'}(I' , R', \mSessKey, S)
        \rightarrow t_1' \doteq t_1)
    \ \ \lor\ \ (\exists t_1\mLogicDot \mRevLTK^{t_1}(R) \land t_1 \lessdot t_2)
\end{align*}
\begin{align*}
    \mPredInjR \triangleq\ &
    \forall I, R, \mSessKey, S, t_2\mLogicDot \mRComplete^{t_2}(I, R, \mSessKey, S)
    \rightarrow\\
    &(\exists t_1\mLogicDot \mIStart^{t_1}(I, R, \mSessKey, S) \land t_1 \lessdot t_2)
    \land (\forall I' R' t_1' \mLogicDot \mRComplete^{t_1'}(I' , R', \mSessKey, S)
        \rightarrow t_1' \doteq t_1)
    \ \ \lor\ \ (\exists t_1\mLogicDot \mRevLTK^{t_1}(I) \land t_1 \lessdot t_2).
\end{align*}
\begin{align*}
    \mPredImpI \triangleq\ &
    \forall I, R, \mSessKey, S, t_1\mLogicDot \mIComplete^{t_1}(I, R, \mSessKey, S)
    \rightarrow\\
      &(\forall I', R', S', t_2\mLogicDot \mRComplete^{t_2}(I', R', \mSessKey, S') \rightarrow
             (I=I' \land R=R' \land S=S'))\\
      &\land (\forall I', R', S', t_1'\mLogicDot
        (\mIComplete^{t_1'}(I', R', \mSessKey, S') \rightarrow t_1' \doteq t_1
        )\\
      &\ \ \ \lor(\exists t_0\mLogicDot \mRevLTK^{t_0}(R) \land t_0 \lessdot t_1)
    \lor(\exists t_0\mLogicDot \mRevEph^{t_0}(R, \mSessKey))
    \lor(\exists t_0\mLogicDot \mRevEph^{t_0}(I, \mSessKey)).
\end{align*}
%
\caption{Formalization of security properties and adversary model.}
\label{fig:props}
\end{figure*}

@@

Having described how we adapt the equational theory to model \mEdhoc,
we now move on to the modelling of the adversary model and the environment in
which the protocol is executed.
%
We extend the built-in Dolev-Yao adversary rules which are part of \mTamarin.
%
We use the following rules to capture the link between a party's identity and
their long-term key pairs, in the \mSig{}- and in the \mStat{}-based methods
respectively.
\begin{center}
\begin{minipage}{0.48\textwidth}
\begin{small}
\begin{verbatim}
1 rule registerLTK_SIG:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_SIG($A, ~ltk),
4         !PK_SIG($A, pk(~ltk)),
5         Out(<$A, pk(~ltk)>)]
\end{verbatim}
\end{small}
\end{minipage}
\hfill\vline\hfill
\begin{minipage}{0.48\textwidth}
\begin{small}
\begin{verbatim}
1 rule registerLTK_STAT:
2    [Fr(~ltk)] --[UniqLTK($A, ~ltk)]->
3        [!LTK_STAT($A, ~ltk),
4         !PK_STAT($A, `g'^~ltk),
5         Out(<$A, `g'^~ltk>)]
\end{verbatim}
\end{small}
\end{minipage}
\end{center}

@@

As we extended the model with TEEs and augmented the adversary's capabilities
with access to them, \mTamarin{} failed to complete in a reasonable time for
some combination of authentication methods and security properties (see Section~\ref{sec:conclusion} for a detailed discussion).
%
To circumvent the problem, we simplified the \mXor{} encryption to \mXor-ing a
single term on the entire tuple for these cases. \\
%

{
\bgroup
\parindent 0pt
\begin{minipage}{0.49\textwidth}
\bgroup
\parindent 0pt
\begin{scriptsize}
\begin{verbatim}
1 rule R2_STAT_SIG:
2 let
3    agreed = <CS0, CI, ~CR>
4    gx = `g'^xx
5    data_2 = <`g'^~yy, CI, ~CR>
6    m1 = <`STAT', `SIG', CS0, CI, gx>
7    TH_2 = h(<$H0, m1, data_2>)
8    prk_2e = extr(`e', gx^~yy)
9    prk_3e2m = prk_2e
10   K_2m = expa(<$cAEAD0, TH_2, `K_2m'>,
11               prk_3e2m)
12   protected2 = $V // ID_CRED_V
13   CRED_V = pkV
14   extAad2 = <TH_2, CRED_V>
15   assocData2 = <protected2, extAad2>
16   MAC_2 = aead(`e', K_2m, assocData2,
                  $cAEAD0)
17   authV = sign(<assocData2, MAC_2>, ~ltk)
18   plainText2 = <$V, authV>
\end{verbatim}
\end{scriptsize}
\egroup
\end{minipage}
\begin{minipage}{0.02\textwidth}
\end{minipage}
\begin{minipage}{0.48\textwidth}
\bgroup
\parindent 0pt
\begin{scriptsize}
\begin{verbatim}
19   K_2e = expa(<$cAEAD0, TH_2,
                 `K_2e'>, prk_2e)
20   K_2e_1 = expa(<$cAEAD0, TH_2,
                   `K_2e', `1'>, prk_2e)
21   K_2e_2 = expa(<$cAEAD0, TH_2,
                   `K_2e', `2'>, prk_2e)
22   CIPHERTEXT_2 = <$V XOR K_2e_1,
                     authV XOR K_2e_2>
23   m2 = <data_2, CIPHERTEXT_2>
24   exp_sk = <gx^~yy>
25 in
26   [!LTK_SIG($V, ~ltk), !PK_SIG($V, pkV),
      In(m1), Fr(~CR), Fr(~yy), Fr(~tid)]
27   --[ExpRunningR(~tid, $V, exp_sk, agreed),
        R2(~tid, $V, m1, m2)]->
28   [StR2_STAT_SIG($V, ~ltk, ~yy, prk_3e2m,
              TH_2, CIPHERTEXT_2, gx^~yy,
              ~tid, m1, m2, agreed),
29    Out(m2)]
\end{verbatim}
\end{scriptsize}
\egroup
\end{minipage}
\\
\egroup
%
}
\vspace{5mm}

@@

The properties we listed in Section~\ref{sec:desired-properties}
translate directly into \mTamarin's logic.
%
We show the \mTamarin{} lemma which encodes the \mPredPfs{} property.
%
Other properties are formalized similarly. \pagebreak
%
